---
title: "A3"
author: "Sebastian Doka"
date: "2023-03-01"
output:
  pdf_document: default
  html_notebook: default
---



Question 1:
A)If I test positive on a Covid Rapid Test, what’s the probability that I actually have Covid?

Let Y represent the probability of having covid, with Y=0 for not having covid and Y=1 for having covid. Let X represent the result of the Covid Rapid Test with X=0 for a negative result and X=1 for a positive result.

P(Y = 1|X = 1) = P(X=1|Y=1)P(Y=1)/P(X=1)

P(Y = 1|X = 1) = P(X=1|Y=1)P(Y=1)/P(X=1|Y=0)*P(Y=0)+P(X=1|Y=1)*P(Y=1)

P(Y = 1|X = 1) = 0.6*0.01/0.005*0.99+0.6*0.01

P(Y = 1|X = 1) = 0.5479452

Therefore the probability of having covid if testing positive for a Covid Rapid Test is 54.79%.

B) If I test negative on a Covid Rapid Test, what’s the probability that I actually have Covid anyway, i.e. the
test was wrong?

P(Y = 1|X = 0) = P(X=0|Y=1)P(Y=1)/P(X=0)

P(Y = 1|X = 0) = P(X=0|Y=1)P(Y=1)/P(X=0|Y=0)*P(Y=0)+P(X=0|Y=1)*P(Y=1)

P(Y = 1|X = 0) = 0.4*0.99/0.995*0.99+0.4*0.01

P(Y = 1|X = 0) = 0.0040443

Therefore the probability of having covid if testing negative for a Covid Rapid Test is 0.4044%.

C) If I test negative twice, and we assume the results from the two tests are independent (a questionable
assumption), then what’s the probability that I actually have Covid?

Because we are assuming the two tests are independent therefore the probability of having covid if negative twice for a Covid Rapid Test is the probabilty of having covid if testing negative once for Covid Rapid squared meaning 0.404^2 = 0.001636%.

Question 2:

A)
```{r echo=FALSE}
library(mnormt)

x <- seq(-5, 5, 0.25)
y <- seq(-5, 5, 0.25)
mu <- c(0,0)
sigma <- matrix(c(1, 0, 0, 4), nrow = 2)
f <- function(x, y) {dmnorm(cbind(x, y), mu, sigma)}
z <- outer(x, y, f)

contour(x, y, z)
```


```{r echo=FALSE}
library(mnormt)

x <- seq(-5, 5, 0.25)
y <- seq(-5, 5, 0.25)
mu <- c(0,0)
sigma <- matrix(c(5, 3, 3, 5), nrow = 2)
f <- function(x, y) {dmnorm(cbind(x, y), mu, sigma)}
z <- outer(x, y, f)

contour(x, y, z)
```

B)
```{r}
library(mnormt)

x <- seq(-5, 5, 0.25)
y <- seq(-5, 5, 0.25)
mu <- c(0,0)
sigma <- matrix(c(3, -1.1, -1.1, 0.9), nrow = 2)
f <- function(x, y) {dmnorm(cbind(x, y), mu, sigma)}
z <- outer(x, y, f)

contour(x, y, z)

```

Question 3

A)
```{r echo=FALSE}
library(palmerpenguins)
df = subset(penguins, species == 'Gentoo')
df = df[,c("sex","body_mass_g","flipper_length_mm","bill_length_mm")]
df = na.omit(df)

library(caret)
set.seed(1)
ind <- createDataPartition(df$sex, p = .5, list=FALSE)
train = df[ind,]
test = df[-ind,]

train_male = subset(train, sex=='male')
train_female = subset(train, sex=='female')

mu_m_flipper = mean(train_male$flipper_length_mm)
sd_m_flipper = sd(train_male$flipper_length_mm)
var_m_flipper = var(train_male$flipper_length_mm)
mu_f_flipper = mean(train_female$flipper_length_mm)
sd_f_flipper = sd(train_female$flipper_length_mm)
var_f_flipper = var(train_female$flipper_length_mm)
  
mu_m_bill = mean(train_male$bill_length_mm)
sd_m_bill = sd(train_male$bill_length_mm)
var_m_bill = var(train_male$bill_length_mm)
mu_f_bill = mean(train_female$bill_length_mm)
sd_f_bill = sd(train_female$bill_length_mm)
var_f_bill = var(train_female$bill_length_mm)

print(paste("Mean of flipper length in subset male:" , round(mu_m_flipper,digits = 4)))
print(paste("Standard deviation of flipper length in subset male:" , sd_m_flipper))
print(paste("Variance of flipper length in subset male:" ,var_m_flipper))
print(paste("Mean of flipper length in subset female:" ,mu_f_flipper))
print(paste("Standard deviation of flipper length in subset female:" ,sd_f_flipper))
print(paste("Variance of flipper length in subset female:" ,var_f_flipper))

print(paste("Mean of bill length in subset male:" ,mu_m_bill))
print(paste("Standard deviation of bill length in subset male:" ,sd_m_bill))
print(paste("Variance of bill length in subset male:" ,var_m_bill))
print(paste("Mean of bill length in subset female:" ,mu_f_bill))
print(paste("Standard deviation of bill length in subset female:" ,sd_f_bill))
print(paste("Variance of bill length in subset female:" ,var_f_bill))

```

B)
```{r}
mu_m = c(mu_m_flipper, mu_m_bill)
mu_f = c(mu_f_flipper, mu_f_bill)
sd_m = c(sd_m_flipper, sd_m_bill)
sd_f = c(sd_f_flipper, sd_f_bill)
# Gaussian Naive Bayes for 2 classes (male and female) and 2 predictor variables
GNB2 = function(X, pi_m, mu_m, sd_m, mu_f, sd_f){
# Inputs:
# X = a dataframe with two columns (the predictors)
# pi_m = a prior probability for class 'male'
# All the other inputs are vectors of length 2, corresponding to 2 predictor variables
# mu_m = a mean vector for class 'male'
# sd_m = a standard deviation vector for class 'male'
# mu_f = a mean vector for class 'female'
# sd_f = a standard deviation vector for class 'female'
##
#Output: a vector of posterior probabilities for class 'male'
for (i in 1:nrow(X)){
likelihood_male = dnorm(X$flipper_length_mm[i], mean=mu_m[1], sd = sd_m[1]) *
dnorm(X$bill_length_mm[i], mean=mu_m[2], sd = sd_m[2])
likelihood_female = dnorm(X$flipper_length_mm[i], mean=mu_f[1], sd = sd_f[1]) *
dnorm(X$bill_length_mm[i], mean=mu_f[2], sd = sd_f[2])
X$posterior_male[i] = (likelihood_male * pi_m) / (likelihood_male * pi_m + likelihood_female * pi_m)
}

return(X$posterior_male)
}

library(class)
predictors = df[,c("flipper_length_mm", "bill_length_mm" )]

grid = expand.grid(x=201:240, y=41:60) 
classes.grid = GNB2(predictors,0.5,mu_m,sd_m,mu_f,sd_f)
classes.grid
```

C)
```{r}
#library(palmerpenguins)
df = subset(penguins, species == 'Gentoo')
predictors = df[,c("flipper_length_mm", "bill_length_mm" )]
#predictors = na.omit(predictors)

mu_m = c(mu_m_flipper, mu_m_bill)
mu_f = c(mu_f_flipper, mu_f_bill)
sd_m = c(sd_m_flipper, sd_m_bill)
sd_f = c(sd_f_flipper, sd_f_bill)

library(class)


grid = expand.grid(x=201:240, y=41:60) 
classes.grid = GNB2(predictors,0.5,mu_m,sd_m,mu_f,sd_f)


#contour(x=201:240, y=41:60, z=matrix(classes.grid, nrow=40), 
 #       levels=0.5,col="grey", drawlabels=FALSE, lwd=2)


#points(df$flipper_length_mm, df$bill_length_mm, col=df$sex)
```

D)
```{r}
library(palmerpenguins)
df = subset(penguins, species == 'Gentoo')
df = df[,c("sex","body_mass_g","flipper_length_mm","bill_length_mm")]
df = na.omit(df)

library(caret)
set.seed(1)
ind <- createDataPartition(df$sex, p = .5, list=FALSE)
train = df[ind,]
test = df[-ind,]

mu_m = c(mu_m_flipper, mu_m_bill)
mu_f = c(mu_f_flipper, mu_f_bill)
var_m = c(mean(var_m_flipper), mean(var_m_bill))
var_f = c(mean(var_f_flipper), mean(var_f_bill))

library(class)
predictors = df[,c("flipper_length_mm", "bill_length_mm" )]

grid = expand.grid(x=201:240, y=41:60) 
classes.grid = GNB2(predictors,0.5,mu_m,var_m,mu_f,var_f)


#contour(x=201:240, y=41:60, z=matrix(classes.grid, nrow=1), levels=0.5,col="grey", drawlabels=FALSE, lwd=2)

 #points(df$flipper_length_mm, df$bill_length_mm, col=df$sex)
```


E)
```{r}

```

